---
title: "Data Structure"
author: "Dr. Xiang Ji @ Tulane University"
date: "Aug 31, 2020"
output:
  html_document:
    toc: true
    toc_depth: 4  
subtitle: MATH-7360 Data Analysis
csl: ../apa.csl
---

```{r}
rm(list = ls()) # clean-up workspace
```

## Announcement
 
 - Lab session
 
 - [HW1](https://github.com/tulane-math7360/tulane-math7360.github.io/raw/master/HW/HW1/HW1_Fall_2020.pdf) posted, due Sep 11
 
 - 1-page report of course project due this week
 
 - [UConn Sports Analytics Symposium](https://statds.org/events/ucsas2020/photos/ucsas2020.pdf), Saturday October 10, $5 registration

## R's data structures

|      | Homogeneous | Heterogeneous |
| ---- | ----------- | ------------- |
| 1d   | Atomic vector | List |
| 2d   | Matrix      | Data frame |
| nd   | Array       |            |

 - _Homogeneous_: all contents must be of the same type
 
 - _Heterogeneous_: the contents can be of different types


## Vectors

 - The basic data structure in R.
 
 - Two flavors: _atomic vectors_ and _lists_
 
 - Three common properties:
 
   - Type, `typeof()`, what it is.
   
   - Length, `length()`, how many elements it contains.
   
   - Attributes, `attributes()`, additional arbitrary metadata.
   
 - No scalars in R.  They are length 1 vectors.

**Note**: `is.vector()` does not test if an object is a vector.  Use `is.atomic()` or `is.list()` to test.

### Atomic vectors

 - There are four common types of atomic vectors (remember Lab 2?)
 
   - logical
   
   - integer
   
   - numeric (actually double)
   
   - character



Many commands in R generate a vector of output, rather than a single number.

The `c()` command: creates a vector containing a list of specific elements.

**Example 1**

```{r}
c(7, 3, 6, 0)
c(73:60)
c(7:3, 6:0)
c(rep(7:3, 6), 0)
```

**Example 2**
The command `seq()` creates a sequence of numbers.
```{r}
seq(7)
seq(3, 70, by = 6)
seq(3, 70, length = 6)
```

 - Atomic vectors are **always flat**, even if you nest `c()`'s:
 
**Example 3**
```{r}
c(1, c(2, c(3, 4)))
```

### Lists

 - Elements can be of any type, including lists.
 
 - Construct list by using `list()` instead of `c()`.
 
```{r}
x <- list(1:3, "a", c(TRUE, FALSE, TRUE), c(2.3, 5.9))
str(x)
```

 - Can be named, can access by name with `$`.
 
```{r}
x.named <- list(vector = 1:3, name = "a", logical = c(TRUE, FALSE, TRUE), range = c(2.3, 5.9))
str(x.named)
x.named$vector
x.named$range
```

 - Lists are used to build up many of the more complicated data structures in R.
 
 - For example, both data frames (another data structure in R) and linear models objects (as produced by `lm()`) are lists.
 
### Attributes
 
 - All objects can have arbitrary additional attributes to store metadata about the object.
 
 - Attributes can be thought as a _named list_.
 
 - Use `attr()` to access individual attribute or `attributes()` to access all attributes as a list.
 
 - By default, most attributes are lost when modifying a vector.  Only the most important ones stay:
 
   - Names, a character vector giving each element a name.
  
   - Dimensions, used to turn vectors into matrices and arrays.
  
   - Class, used to implement [S3](http://adv-r.had.co.nz/OO-essentials.html#s3) object system.
 
```{r}
y <- 1:10
attr(y, "my_attribute") <- "This is a vector"
attr(y, "my_attribute")
str(y)
str(attributes(y))
```

### Factors

 - A factor is a vector that can contain only predefined values and is used to store _categorical_ data.
 
 - Built upon integer vectors using two attributes: 
   
   - the `class`, "factor": makes them behave differently from regular integer vectors
   
   - the `levels`: defines the set of allowed values
   
 - Sometimes when a data frame is read directly from a file, you may get a column of factor instead of numeric because of non-numeric value in the column (e.g. missing value encoded specially)
 
   - Possible remedy: coerce the vector from a factor to a character vecctor, and then from a character to a double vector
   
   - Better use `na.strings` argument to `read.csv()` function

### Operations on vectors

Use brackets to select element of a vector.

```{r}
x <- 73:60
x[2]
x[2:5]
x[-(2:5)]
```

Can access by "name" (safe with column/row order changes)

```{r}
y <- 1:3
names(y) <- c("do", "re", "mi")
y[3]
y["mi"]
```

### Matrices and arrays

 - adding a `dim` attribute to an atomic vector allows it to behave like a multi-dimensional **array**
 
 - **matrix** is a special case of array

 - `matrix()` command creates a matrix from the given set of values

```{r}
# Two scalar arguments to specify rows and columns
a <- matrix(1:6, ncol = 3, nrow = 2)
# One vector argument to describe all dimensions
b <- array(1:12, c(2, 3, 2))

# You can also modify an object in place by setting dim()
c <- 1:6
dim(c) <- c(3, 2)
c
dim(c) <- c(2, 3)
c
```


### R commands on vector/matrix

| command | usage |
|---------|-------|
| sum()   | sum over elements in vector/matrix |
| mean()  | compute average value |
|sort() | sort all elements in a vector/matrix |
| min(), max() | min and max values of a vector/matrix|
| length() | length of a vector/matrix |
|summary() | returns the min, Q1, median, mean, Q3, and max values of a vector |
| dim() | dimension of a matrix |
| cbind() | combine a sequence of vector, matrix or data-frame arguments and combine by columns |
| rbind() | combine a sequence of vector, matrix or data-frame arguments and combine by rows |
| names() | get or set names of an object |
| colnames() | get or set column names of a matrix-like object |
| rownames() | get or set row names of a matrix-like object |
<!-- | sample(x, size, replace = FALSE, prob = NULL) | takes a random sample from a vector with or without replacement| -->

**Exercise**
Write a command to generate a random permutation of the numbers between 1 and 5 and save it to an object.

### Data frames

 - Most common way of storing data in R
 
 - A list of equal-length vectors
 
 - 2-dimensional structure, shares properties of both `matrix` and `list`
 
   - has attributes, `names()`, `colnames()` and `rownames()`
   
   - `length()` of a data frame is the length of the underlying list, same as `ncol()`
   
 - More in this week's lab session


<!--
```{r}
order(runif(5))
sample(1:5, 5)
```
-->




